---
title: "Functional Programming - Back to Basics"
author: "Eryn McFarlane"
date: "09/02/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Why do we do Functional Programming
(this is very loosely based off https://ourcodingclub.github.io/tutorials/funandloops/, among others)

The whole point is to 1) make your life easier because you don't need to manually re-do the same processes you've already done and 2) cut down on error by copy and pasting. 

General rule: if you're going to do something more than twice, it's worth figuring out a smarter way to do it. 

For example, you might want to do multiple plots, which differ only in their input (e.g. graphing the same x and y for different years). You might want to process data from different files in the same way etc. 

If you write a function or a loop, then you will only need to change code in one place, if you need to change it, as compared to finding all of the places you might have implemented it (again, increasing the likelihood of errors). 


```{r}
### make up some data
df<-data.frame(cbind(rnorm(100, 0, 1), rnorm(100, 5, 3)))
```

Let's write a (very simple) function:

```{r}
function1<-function(x,y){
  #do something with x and y
  x+y
}

function1(df[,1], df[,2])

head(df[,1]+df[,2])
```

Obviously this is a really simple function. I've just told the function that I was going to give it two arguments, and that I wanted it to add them together. 

Now, we don't need to only use this on the made up data, we can use it on any two arguments

```{r}

function1(1, 100)
```

We can make this function as simple or as difficult as we like. If it's something that is going to be used accross multiple projects, it's worth saving it separately so that it can be called using source(). 

We can also put as many arguments into as we like. We can also name them whatever we like. 

```{r}
function2<-function(one, two, red, blue){
  (one+two)*red/blue
}

function2(1, 2, 3, 4)
```
## For Loops
Let's look a little bit at loops. 

In R, there are for loops and while loops. I've never used a while loop, so for loops. 

For loops iterate through the items, and perform some action on each of them. 


```{r}
for(i in 1:3){
  ## do the thing
  print("the thing")
}
```

Perhaps with our data, we want to standardize each point by the mean and the standard deviation. 

```{r}

for(i in 1:length(df[,1])){
  df[i, 3]<-df[i,1]-mean(df[,1])/sd(df[,1])
}
```

Excellent. Except maybe we want to do this for each of our rows.


```{r}
str(df)
df<-df[,c(1:2)] # get back to original df
df_std<-matrix(NA, nrow=100, ncol=2) ##make a new matrix to chuck the standardize values in
#### this is something that always trips me up with loops, I don't find it intuitive to catch the data. 
### making a list, matrix or vector before the loop is the way to go. 
## alternatively, there is a package called 'magicfor' https://cran.r-project.org/web/packages/magicfor/vignettes/magicfor.html
#which can catch data coming out of loops if you start it before your loop. I've used this for monster simulations and, at the time, it felt like it was making my life easier, but I haven't used it in ages, so YMMV

for(j in 1:ncol(df)){
  for(i in 1:nrow(df)){
  df_std[i, j]<-(df[i,j]-mean(df[,j]))/sd(df[,j])
    }
}

head(df_std)
```

So you can see that you can put loops inside loops inside loops. This is not always efficient, and it's easy to lose some things. Additionally, the way that loops are saved can slow them down a lot, which can be annoying if there's a lot of data, or many iterations. 

This is where the amazing 'apply' group of functions comes in.


For example, we can write a function, and run it on df using lappy. 
Let's take our fake data, and add a category to it. 

```{r}
df[,3]<-rep(c("yes", "no"), 50)
function3<-function(x){
  (x-mean(x))/sd(x)
}

lapply(df[,c(1:2)], function3)
head(sapply(df[,c(1:2)], function3), 10)
tapply(df[,1], df[,3], function3)

```
The first argument of lapply() gives the list object to be iterated, and the second object is the function. 
sapply() simplfies lapply() to get the output into a vector.
tapply() allows you to run an apply with a different condition. 

Additionally, you can build apply functions into other things. 

```{r}
df[,4]<-rep(c(1:50), 2) ### this can be the unique identifier
df[,5]<-sample(rep(c("one", "two", "red", "blue"), 25))
with(df, tapply(df[,4], list(df[,5], df[,3]), function(x) length(unique(x))))
```

Finally, while I've written function3, it already exists! This is what scale() does. Sometimes cool to reinvent the wheel, sometimes not needed. 

```{r}
head(scale(df[,1]))
head(scale(df[,2]))
```

Of course this is all do-able in tidyverse as well. I've not gone into it here, but I think it's the map group of functions (I only use tidyverse when absolutely necessary)


## Conditional Statements

For example, an ifelse() statement. Based on a condition, do this, or do this. These can be really simple, or can get really complicated, and can also be nested.
Of course, this could then be put into a for loop, or into an apply. 

```{r}
newvariable<-ifelse(df[,5]=='blue', df[,1]+6, df[,1])

nested_conditional<-ifelse(df[,5]=='blue', df[,1]*4, 
                           ifelse(df[,5]=="red", df[,1]/5, 
                           ifelse(df[,5]=="one", df[,1]+7, 
                           df[,1])))

head(nested_conditional)

```